Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die Einhaltung der Echtzeit

### 2024-08-23 Masterarbeit_Robotik_Pamuk.tex
3.3 Real-Time Robotic Application, 3.3.1 Setup of Hardware Salamander 4, 3.3.2 Setup of QEMU Salamander 4

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=blue!20,
					text width=35em,
					rounded corners,
					minimum height=3.5em,
					align=center,
					inner sep=0.2em % Adjusted inner separation
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (init) {\scriptsize \textbf{Guest Code}\\  The guest OS is executing its normal operations.};
		\node [block, below=of init] (vmexit) {\scriptsize \textbf{VM Exit}\\  External interrupt triggers VM exit. Hardware stores interrupt info in VMCS.};
		\node [block, below=of vmexit] (aftervmexit) {\scriptsize \textbf{After VM Exit}\\  Control returns to kernel space. KVM retrieves interrupt info from VMCS,\\\vspace{-0.5em} marks it as pending for guest.};
		\node [block, below=of aftervmexit] (hostisr) {\scriptsize \textbf{Host ISR}\\  Host Linux handles the interrupt.};
		\node [block, below=of hostisr] (hostscheduling) {\scriptsize \textbf{Host Scheduling}\\  Host kernel invokes scheduler. Ideally, guest RTOS process isn't preempted.};
		\node [block, below=of hostscheduling] (beforevmentry) {\scriptsize \textbf{Before VM Entry}\\  KVM's kernel module prepares for VM entry. Detects pending interrupt,\\\vspace{-0.5em} injects virtual interrupt into guest by writing to VMCS.};
		\node [block, below=of beforevmentry] (vmentry) {\scriptsize \textbf{VM Entry}\\  Hardware emulates conventional interrupt handling for guest based on VMCS info.};
		\node [block, below=of vmentry] (guestirt) {\scriptsize \textbf{Guest IRT}\\  Guest's interrupt handler is invoked to service the interrupt.};
		% Draw edges
		\path [line] (init) -- (vmexit);
		\path [line] (vmexit) -- (aftervmexit);
		\path [line] (aftervmexit) -- (hostisr);
		\path [line] (hostisr) -- (hostscheduling);
		\path [line] (hostscheduling) -- (beforevmentry);
		\path [line] (beforevmentry) -- (vmentry);
		\path [line] (vmentry) -- (guestirt);
	\end{tikzpicture}
	\caption{Flowchart of operations in a virtualized environment (own figure, inspired by~\citeauthor{maPerformanceTuningKVMbased})}
	\label{fig:kvm_stages}
\end{figure}

@@

\section{Kernelshark}\label{sec:kernelshark}
Viewing the tracing data that was recorded with \texttt{trace-cmd} on a text-based file can be laborious. A better way of analyzing the data is done through \texttt{KernelShark}~\cite{KernelShark}, which is a graphical front-end tool. It visualizes the recorded kernel trace data in a readable way on an interactive timeline, which facilitates the process of identifying patterns and correlations between events. The latency issues can be analyzed by further filtering the displayed events according to processes, event types or time ranges. After the recording, \texttt{trace-cmd} gernerates a host \texttt{trace.dat} file and a guest \texttt{trace-Salamander4.dat} file, which in this case can be loaded into \texttt{KernelShark} with the command \texttt{kernelshark trace.dat -a trace-Salamander4.dat}. Once both files are loaded, the trace data can be visually inspected. All the current CPUs for both the host and the guest can be turned off, and the KVM Combo plots menu can be selected to view the guest vCPUs plotted on top of the host threads that run them, as depicted in Figure~\ref{fig:kernelshark_combo}.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/kernelshark_combo.png}
	\caption[Guest vCPUs plotted on top of host threads in KernelShark]{Guest vCPUs plotted on top of host threads in KernelShark}
	\label{fig:kernelshark_combo}
\end{figure}

@@

\section{Real-Time Robotic Application}\label{sec:robotic_application}
This chapter compares the latency of the Salamander 4 operating system before and after the real-time performance tunings with the latency of Salamander 4 running on bare metal hardware. This comparison is done with the aid of a program that will be explained shortly. The goal is to understand how closely the performance of the virtualization can match that of the bare metal. The experimental setup includes a six-axis mini-robot, illustrated in Figure~\ref{fig:mini_robot}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\columnwidth]{img/experiment/mini_robot.jpg}
	\caption[Mini-robot of the experiment]{Mini-robot of the experiment}
	\label{fig:mini_robot}
\end{figure}

\noindent The drive system of the robot arm consists of six MG996R digital servo motors~\cite{MG996RServoMotor}, equipped with a metal gearbox. The motor is able to rotate in a range of approximately 180 degrees and its position can be controlled with a high degree of accuracy. Each servo motor has three wires that need to be connected as shown in Figure~\ref{fig:motor_wires}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\columnwidth]{img/experiment/motor_wires.png}
	\caption[MG996R Servo Motor]{MG996R Servo Motor~\cite{MG996RServoMotor}}
	\label{fig:motor_wires}
\end{figure}

\noindent To drive the motor, it has to be powered using the red and brown wires and can be controlled by sending PWM signals to the orange wire. In this experiment, this PWM signal is generated by the proprietary PW 022 pulse width module of Sigmatek~\cite{DigitalOutputSIGMATEK}. Its connector layout is illustrated in Figure~\ref{fig:pw022_connectors}.

@@

\noindent The module has two +24 V switching PWM outputs with an adjustable frequency for controlling inductive loads. Since the mentioned servo motors operate between 4.8 volts and 7.2 volts~\cite{MG996RServoMotor}, this voltage needed to be reduced through resistors before supplying it to a servo motor. For this purpose, two resistors with resistances of 2500 kiloohms and 1000 kiloohms were connected in series. The voltage across each resistor is calculated using Ohm's law, which is given by equation~\ref{eq:ohms_law} below.
\begin{equation}
	\label{eq:ohms_law}
	V = I \cdot R
\end{equation}
In this equation, \(V\) represents the voltage across the resistor, \(I\) is the current flowing through the resistor, and \(R\) is the resistance of the resistor. The resulting voltages across the resistors are as follows:
\begin{itemize}
	\item The voltage across the 2.5 kiloohm resistor is approximately 17.14 volts.
	\item The voltage across the 1 kiloohm resistor is approximately 6.86 volts. This voltage was then supplied to the control wire of servo motor 1 of the mini-robot.
\end{itemize}

@@

\noindent The program was written in Lasal Class 2 and was applied to all three mentioned versions of Salamander 4 to measure the reaction time of the robot to the specific commands. Prior to examining the software program, the next two subsections briefly explain the setup of each version of the experiment.

\subsection{Setup of Hardware Salamander 4}
In this version, Salamander 4 runs on the CP 841~\cite{CPUEinheitenSIGMATEK} CPU unit, specifically designed for the Salamander 4 operating system. The PW 022 module is directly mounted on the CPU via the S-DIAS bus and communicates over the hard real-time capable Ethernet VARAN with 100 Mbit/s~\cite{SDIASSIGMATEK}. The setup is visible on Figure~\ref{fig:hardware_tree}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\columnwidth]{img/experiment/hardware_tree.png}
	\caption[Hardware tree]{Hardware tree}
	\label{fig:hardware_tree}
\end{figure}

\subsection{Setup of QEMU Salamander 4}\label{subsec:setup_experiment_virtualized}
In the virtualized setup of the experiment, the VCPU functionality of QEMU is used to connect QEMU with the PWM module. In order to achieve that, the PCV 522 VARAN Manager PCI Insert Card~\cite{ControlsHMIsSIGMATEK}, which serves as a bridge between the PC and the rest of the setup, needed to be plugged into the PC. The command \texttt{lspci -nn} lists all PCI devices along with their vendor and device IDs. In this case, the command for binding the PCV 522 module to the VFIO-PCI driver was \texttt{sudo sh -c 'echo "5112 2200" > /sys/bus/pci/drivers/vfio-pci/new\_id'} with vendor ID 5112 and device ID 2200.  To verify that the device has been successfully bound to the VFIO-PCI driver, \texttt{lspci -v} can be used. On top of the binding process, the QEMU script from the previous sections also needed to be modified to include the PCV 522 VARAN Manager PCI Insert Card. This is done by adding \texttt{-device\ vfio-pci,host=03:00.0 } to the QEMU script, where \texttt{03:00.0} refers to the PCI address of the device, with \texttt{03} being the bus number, \texttt{00} the device number, and \texttt{0} the function number. The final script can be seen in Code~\ref{script:qemu_def_pci}.

\vspace{2em}
\begin{lstlisting}[name={Include PCI in QEMU script for Salamander 4 virtualization},label={script:qemu_def_pci}]
	#!/bin/sh

@@

	exec taskset -c 4-5 chrt -f 99 qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
	-m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
	-append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4 schedstats=enable nohlt idle=poll quiet xeno_hal.smi=1 xenomai.smi=1 threadirqs" \
	-net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
	-fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
	-device vhost-vsock-pci,guest-cid=3,id=vsock0 \
	-drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
	-object memory-backend-ram,id=ram0,size=4G,prealloc=on \
	-mem-prealloc -mem-path /dev/hugepages \
	-device vfio-pci,host=03:00.0 \
	-no-reboot -nographic
\end{lstlisting}

\noindent An additional Varan Connection module VI 021~\cite{InterfacesSplittersSIGMATEK} was required to enable the connection between the PCV 522 module and the PW 022 module that generates the signal to move the mini-robot. This setup is depicted in Figure~\ref{fig:virt_tree}.

@@

\noindent Essentially, the CP 841 CPU is being emulated by the VCPU functionality of QEMU, allowing the PCV 522 VARAN Manager PCI Insert Card and the Varan Connection module VI 021 to interact as if they were communicating with a physical CPU unit.

\clearpage
\subsection{Robotic Application}

>>>>

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=blue!20,
					text width=35em,
					rounded corners,
					minimum height=3.5em,
					align=center,
					inner sep=0.2em % Adjusted inner separation
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (init) {\scriptsize \textbf{Guest Code}\\  The guest OS is executing its normal operations.};
		\node [block, below=of init] (vmexit) {\scriptsize \textbf{VM Exit}\\  External interrupt triggers VM exit. Hardware stores interrupt info in VMCS.};
		\node [block, below=of vmexit] (aftervmexit) {\scriptsize \textbf{After VM Exit}\\  Control returns to kernel space. KVM retrieves interrupt info from VMCS,\\\vspace{-0.5em} marks it as pending for guest.};
		\node [block, below=of aftervmexit] (hostisr) {\scriptsize \textbf{Host ISR}\\  Host Linux handles the interrupt.};
		\node [block, below=of hostisr] (hostscheduling) {\scriptsize \textbf{Host Scheduling}\\  Host kernel invokes scheduler. Ideally, guest RTOS process isn't preempted.};
		\node [block, below=of hostscheduling] (beforevmentry) {\scriptsize \textbf{Before VM Entry}\\  KVM's kernel module prepares for VM entry. Detects pending interrupt,\\\vspace{-0.5em} injects virtual interrupt into guest by writing to VMCS.};
		\node [block, below=of beforevmentry] (vmentry) {\scriptsize \textbf{VM Entry}\\  Hardware emulates conventional interrupt handling for guest based on VMCS info.};
		\node [block, below=of vmentry] (guestirt) {\scriptsize \textbf{Guest IRT}\\  Guest's interrupt handler is invoked to service the interrupt.};
		% Draw edges
		\path [line] (init) -- (vmexit);
		\path [line] (vmexit) -- (aftervmexit);
		\path [line] (aftervmexit) -- (hostisr);
		\path [line] (hostisr) -- (hostscheduling);
		\path [line] (hostscheduling) -- (beforevmentry);
		\path [line] (beforevmentry) -- (vmentry);
		\path [line] (vmentry) -- (guestirt);
	\end{tikzpicture}
	\caption[Flowchart of operations in a virtualized environment]{Flowchart of operations in a virtualized environment (own figure, inspired by~\citeauthor{maPerformanceTuningKVMbased})}
	\label{fig:kvm_stages}
\end{figure}

@@

\section{Kernelshark}\label{sec:kernelshark}
Viewing the tracing data that was recorded with \texttt{trace-cmd} on a text-based file can be laborious. A better way of analyzing the data is done through \texttt{KernelShark}~\cite{KernelShark}, which is a graphical front-end tool. It visualizes the recorded kernel trace data in a readable way on an interactive timeline, which facilitates the process of identifying patterns and correlations between events. The latency issues can be analyzed by further filtering the displayed events according to processes, event types or time ranges. After the recording, \texttt{trace-cmd} gernerates a host \texttt{trace.dat} file and a guest \texttt{trace-Salamander4.dat} file, which in this case can be loaded into \texttt{KernelShark} with the command \texttt{kernelshark trace.dat -a trace-Salamander4.dat}. Once both files are loaded, the trace data can be visually inspected. All the current CPUs for both the host and the guest can be turned off, and the KVM Combo plots menu can be selected to view the guest VCPUs plotted on top of the host threads that run them, as depicted in Figure~\ref{fig:kernelshark_combo}.

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/kernelshark_combo.png}
	\caption[Guest VCPUs plotted on top of host threads in KernelShark]{Guest VCPUs plotted on top of host threads in KernelShark}
	\label{fig:kernelshark_combo}
\end{figure}

@@

\section{Real-Time Robotic Application}\label{sec:robotic_application}
This chapter compares the latency of the Salamander 4 operating system before and after the real-time performance tunings with the latency of Salamander 4 running on bare metal. This comparison is done with the aid of a program that will be explained shortly. The goal is to understand how closely the performance of the virtualization can match that of the bare metal. The experimental setup includes a six-axis mini-robot, illustrated in Figure~\ref{fig:mini_robot}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\columnwidth]{img/experiment/mini_robot.jpg}
	\caption[Mini-robot of the experiment]{Mini-robot of the experiment~\cite{6DFRoboticArm}}
	\label{fig:mini_robot}
\end{figure}

\noindent The robot arm consists of six MG996R digital servo motors~\cite{MG996RServoMotor}, equipped with a metal gearbox. The motor is able to rotate in a range of approximately 180 degrees and its position can be controlled with a high degree of accuracy. Each servo motor has three wires that need to be connected as shown in Figure~\ref{fig:servo}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/experiment/servo.jpg}
	\caption[MG996R Servo Motor]{MG996R Servo Motor~\cite{PackMg996rMetal}}
	\label{fig:servo}
\end{figure}

\noindent To drive the motor, it has to be powered using the red and brown wires and can be controlled by sending PWM signals to the orange wire. A 1,500 microsecond pulse every 20 milliseconds centers the servo. A 1,000 microsecond pulse turns it 90 degrees left, and a 2,000 microsecond pulse turns it 90 degrees right. This is depicted in Figure~\ref{fig:servo_angle}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/experiment/servo_angle.png}
	\caption[Controlling the MG996R Servo Motor]{Controlling the MG996R Servo Motor~\cite{twierengEmilyQuadrupedDog}}
	\label{fig:servo_angle}
\end{figure}

\noindent In this experiment, this PWM signal is generated by the proprietary PW 022 pulse width module of Sigmatek~\cite{DigitalOutputSIGMATEK}. Its connector layout is illustrated in Figure~\ref{fig:pw022_connectors}.

@@

\noindent The module has two +24 V switching PWM outputs with an adjustable frequency for controlling inductive loads. Since the mentioned servo motors operate between 4.8 volts and 7.2 volts~\cite{MG996RServoMotor}, this voltage needs to be reduced through resistors before supplying it to a servo motor. For this purpose, two resistors with resistances of 2500 kiloohms and 1000 kiloohms are connected in series. The voltage across each resistor is calculated using Ohm's law, which is given by equation~\ref{eq:ohms_law} below.
\begin{equation}
	\label{eq:ohms_law}
	V = I \cdot R
\end{equation}

\noindent In this equation, \(V\) represents the voltage across the resistor, \(I\) is the current flowing through the resistor, and \(R\) is the resistance of the resistor. The resulting voltages across the resistors are as follows:

\begin{itemize}
	\item The voltage across the 2.5 kiloohm resistor is approximately 17.14 volts.
	\item The voltage across the 1 kiloohm resistor is approximately 6.86 volts. This voltage is then supplied to the control wire of servo motor 1 of the mini-robot.
\end{itemize}

@@

\noindent The program is written in Lasal Class 2 and is applied to all three mentioned versions of Salamander 4 to measure the reaction time of the robot to the commands. Prior to examining the software program, the next two subsections briefly explain the setup of each version of the experiment.

\subsection{Setup of Hardware Salamander 4}
In this version, Salamander 4 runs on the CP 841~\cite{CPUEinheitenSIGMATEK} CPU unit, specifically designed for the Salamander 4 operating system. The PW 022 modules are directly mounted on the CPU via the S-DIAS bus~\cite{SDIASSIGMATEK}. There are three PW 022 modules because each can control two motors simultaneously, and the mini-robot has six motors. The setup is visible on Figure~\ref{fig:hardware_tree}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.45\columnwidth]{img/experiment/hardware_tree.png}
	\caption[Setup of Hardware Salamander 4]{Setup of Hardware Salamander 4}
	\label{fig:hardware_tree}
\end{figure}

\subsection{Setup of QEMU Salamander 4}\label{subsec:setup_experiment_virtualized}
In the virtualized setup of the experiment, the VCPU functionality of QEMU is used to connect QEMU with the PWM modules. In order to achieve that, the PCV 522 VARAN Manager PCI Insert Card~\cite{ControlsHMIsSIGMATEK}, which serves as a bridge between the PC and the rest of the setup, needs to be plugged into the PC. The command \texttt{lspci -nn} lists all PCI devices along with their vendor and device IDs. In this case, the command for binding the PCV 522 module to the VFIO-PCI driver was \texttt{sudo sh -c 'echo "5112 2200" > /sys/bus/pci/drivers/vfio-pci/new\_id'} with vendor ID \texttt{5112} and device ID \texttt{2200}.  To verify that the device has been successfully bound to the VFIO-PCI driver, \texttt{lspci -v} can be used. On top of the binding process, the QEMU script from the previous sections also needed to be modified to include the PCV 522 VARAN Manager PCI Insert Card in the virtualization. This is done by adding \texttt{-device\ vfio-pci,host=03:00.0 } to the QEMU script, where \texttt{03:00.0} refers to the PCI address of the device, with \texttt{03} being the bus number, \texttt{00} the device number, and \texttt{0} the function number. The final script can be seen in Code~\ref{script:qemu_def_pci}.

\clearpage
\vspace*{-5em}
\begin{lstlisting}[name={Include PCI in QEMU for Salamander 4 Virtualization},label={script:qemu_def_pci}]
		#!/bin/sh

@@

		exec taskset -c 4 chrt -f 99 qemu-system-x86_64 -M pc,accel=kvm -kernel ./bzImage \
		-m 2048 -drive file=salamander-image-sigmatek-core2.ext4,format=raw,media=disk \
		-append "console=ttyS0 console=tty1 root=/dev/sda rw panic=1 sigmatek_lrt.QEMU=1 ip=dhcp rootfstype=ext4 schedstats=enable nohlt idle=poll quiet xeno_hal.smi=1 xenomai.smi=1 threadirqs" \
		-net nic,model=e1000,netdev=e1000 -netdev bridge,id=e1000,br=nm-bridge \
		-fsdev local,security_model=none,id=fsdev0,path=drive-c -device virtio-9p-pci,id=fs0,fsdev=fsdev0,mount_tag=/mnt/drive-C \
		-device vhost-vsock-pci,guest-cid=3,id=vsock0 \
		-drive if=pflash,format=qcow2,file=ovmf.code.qcow2 \
		-object memory-backend-ram,id=ram0,size=4G,prealloc=on \
		-mem-prealloc -mem-path /dev/hugepages \
		-device vfio-pci,host=03:00.0 \
		-no-reboot -nographic
	\end{lstlisting}
\vspace{2em}
\noindent An additional Varan Connection module VI 021~\cite{InterfacesSplittersSIGMATEK} was required to enable the connection between the PCV 522 module and the PW 022 module that generates the signal to move the mini-robot. This setup is depicted in Figure~\ref{fig:virt_tree}.

@@

\noindent Essentially, the CPU is being emulated by the VCPU functionality of QEMU, allowing the PCV 522 VARAN Manager PCI Insert Card and the Varan Connection module VI 021 to interact as if they were communicating with a physical CPU unit.

\subsection{Robotic Application}
